Amplicon sequence variants artificially split bacterial genomes into separate clusters

### 2021-06-23 submission/manuscript.Rmd
Respond to reviewers' comments

16S rRNA gene sequencing is a powerful technique for describing and comparing microbial communities [@Lane1985]. Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds date to at least the 1990s. The distance-based threshold that was developed and is now widely used was based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007]. Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018]. As an alternative to operational taxonomic units (OTUs), amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014]. Approaches for identifying ASVs do not cluster sequences based on a distance-based threshold [@Callahan2017]. Proponents of ASVs are largely dissmissive of concerns that most bacterial genomes have more than one copy of the *rrn* operon and that those copies are not identical [@Pei2010; @Sun2013]. Yet, ASVs and using too fine a threshold to identify OTUs could split a single genome into multiple clusters. Conversely, using too broad of a threshold to define OTUs could cluster together multiple bacterial species into the same OTU. An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S. epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. Each of the 10 copies of the 16S rRNA gene in these two genomes is distinct and represent 10 ASVs. Conversely, if the copies were clustered using a 3% distance threshold, then all 10 ASVs would cluster into the same OTU. The goal of this study was to quantify the tradeoff of splitting a single genome into multiple clusters and the risk of clustering different bacterial species into the same cluster when using ASVs and various OTU definitions.

@@

At broad thresholds, 16S rRNA gene sequences from multiple species could be clustered into the same ASV or OTU. I again randomly selected one genome from each species to control for uneven representation of genomes across species and for this analysis I measured the percentage of ASVs and OTUs that contained 16S rRNA gene sequences from multiple species (Figure 2). Without using distance-based thresholds, `r format_pct(lump_v19)`% of the ASVs contained sequences from multiple species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively. At the commonly used 3% threshold for defining OTUs, `r format_pct(lump3_v19)`% of the OTUs contained 16S rRNA gene sequences from multiple species when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively. Considering that species designations are inconsistently applied and reflect multiple human-imposed biases, the risk of splitting a genome into multiple OTUs is more problematic than clustering species together. Therefore, larger thresholds are advisable.

The results of this analysis demonstrate that there is a significant risk of splitting a single genome into multiple clusters if using ASVs or too fine of a threshold to define OTUs. An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit [@Callahan2017; @Schloss2011; @Johnson2019]. Since there is no consensus for a biologicaly definition of a bacterial species [@Staley2006; @Oren2013; @Sanford2021], microbiologists must accept that how bacterial species are named is biased and that taxonomic rules are not applied in a consistent manner (e.g., [@Baltrus2016; @Sanford2021]). This makes it impossible to fit a distance threshold to define an OTU definition that matches a set of species names [@Konstantinidis2005]. Furthermore, the 16S rRNA gene does not evolve at the same rate across all bacterial lineages [@Schloss2011], which limits the biological interpretation of a common OTU definition. A distance-based definition of a taxonomic unit based on 16S rRNA gene or full genome sequences is, at best, operational and not grounded in biological theory [@Schloss2011; @Barco2020; @Parks2020; @Yarza2014]. There is general agreement in bacterial systematics that to classify an organism to a bacterial species, phenotypic and genome sequence data are needed [@Staley2006; @Oren2013; @Baltrus2016; @Sanford2021]. A short sequence from a bacterial genome simply cannot differentiate between species. Moreover, it is difficult to defend a clustering threshold that would split a single genome into multiple taxonomic units. It is not biologically plausible to entertain the possibility that different *rrn* operons from the same genome would have different ecologies. Although there are multiple reasons that proponents favor ASVs, the significant risk of artificially splitting genomes into separate clusters is too high to warrant their use.

@@

\includegraphics{../figures/copy_number_threshold_plot.pdf}

@@

\includegraphics{../figures/lump_split.pdf}

@@

\includegraphics{../figures/esv_rate.pdf}

>>>>

16S rRNA gene sequencing is a powerful technique for describing and comparing microbial communities [@Lane1985]. Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds date to at least the 1990s. The distance-based threshold that was developed and is now widely used was based on DNA-DNA hybridization approaches that are not as precise as genome sequencing [@Stackebrandt1994; @Goris2007]. Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse [@RodriguezR2018; @Stackebrandt2006; @Edgar2018]. As an alternative to operational taxonomic units (OTUs), amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences [@Edgar2016; @Amir2017; @Callahan2016; @Eren2014]. It is widely understood that individual bacterial genomes often have multiple 16S rRNA genes that are not identical and that a 16S rRNA gene sequence could be found with different versions of the sequence in different genomes [@Pei2010; @Sun2013]. This could lead to the problem that ASVs and using too fine a threshold to identify OTUs could split a single genome into multiple clusters. Proponents of ASVs minimize concerns that most bacterial genomes have more than one copy of the *rrn* operon and that those copies are not identical [@Callahan2017; @Edgar2018]. Conversely, using too broad of a threshold to define OTUs could cluster together multiple bacterial species into the same OTU. An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S. epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. Each of the 10 copies of the 16S rRNA gene in these two genomes is distinct and represent 10 ASVs. Conversely, if the copies were clustered using a 3% distance threshold, then all 10 ASVs would cluster into the same OTU. The goal of this study was to quantify the tradeoff of splitting a single genome into multiple clusters and the risk of clustering different bacterial species into the same cluster when using ASVs and various OTU definitions.

@@

At broad thresholds, 16S rRNA gene sequences from multiple species could be clustered into the same ASV or OTU. I again randomly selected one genome from each species to control for uneven representation of genomes across species. For this analysis I measured the percentage of ASVs and OTUs that contained 16S rRNA gene sequences from multiple species (Figure 2). Without using distance-based thresholds, `r format_pct(lump_v19)`% of the ASVs contained sequences from multiple species when considering full length sequences and `r format_pct(lump_v34)`, `r format_pct(lump_v4)`, and `r format_pct(lump_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively. At the commonly used 3% threshold for defining OTUs, `r format_pct(lump3_v19)`% of the OTUs contained 16S rRNA gene sequences from multiple species when considering full length sequences and `r format_pct(lump3_v34)`, `r format_pct(lump3_v4)`, and `r format_pct(lump3_v45)`% when considering the V3-V4, V4, and V4-V5 regions, respectively. Although the actual fraction of ASVs and OTUs that contain sequences from multiple species is dependent on the taxonomic composition of the sequences in the *rrn*DB, this analysis highlights the tradeoffs of using distance-based thresholds.

The results of this analysis demonstrate that there is a significant risk of splitting a single genome into multiple clusters if using ASVs or too fine of a threshold to define OTUs. An ongoing problem for amplicon-based studies is defining a meaningful taxonomic unit [@Callahan2017; @Schloss2011; @Johnson2019]. Since there is no consensus for a biologicaly definition of a bacterial species [@Staley2006; @Oren2013; @Sanford2021], microbiologists must accept that how bacterial species are named is biased and that taxonomic rules are not applied in a consistent manner (e.g., [@Baltrus2016; @Sanford2021]). This makes it impossible to fit a distance threshold to define an OTU definition that matches a set of species names [@Konstantinidis2005]. Furthermore, the 16S rRNA gene does not evolve at the same rate across all bacterial lineages [@Schloss2011], which limits the biological interpretation of a common OTU definition. A distance-based definition of a taxonomic unit based on 16S rRNA gene or full genome sequences is operational and not necessarily grounded in biological theory [@Schloss2011; @Barco2020; @Parks2020; @Yarza2014]. One benefit of a distance-based OTU definition is the ability to mask residual sequencing error. The analysis in this study was conducted using ideal sequences from assembled genomes whereas sequences generated in microbiome studies would harbor PCR and sequencing errors. These errors would only exacerbate the inflated number of ASVs. There is general agreement in bacterial systematics that to classify an organism to a bacterial species, phenotypic and genome sequence data are needed [@Staley2006; @Oren2013; @Baltrus2016; @Sanford2021]. A short sequence from a bacterial genome simply cannot differentiate between species. Moreover, it is difficult to defend a clustering threshold that would split a single genome into multiple taxonomic units. It is not biologically plausible to entertain the possibility that different *rrn* operons from the same genome would have different ecologies. Individual bacteria are defined at the cellular or chromosomal level and not at the gene level. One could argue that, in practice, communities are compared on a relative rather than absolute basis. However, communities harboring populations that tend to have more copies of the *rrn* operon would appear to have higher richness and diversity than those with fewer copies purely due to the propensity for populations with more *rrn* operons to generate more ASVs. Although there are multiple reasons that proponents favor ASVs, the significant risk of artificially splitting genomes into separate clusters is too high to warrant their use.

@@

@@

@@

### 2021-06-23 submission/manuscript.tex
Respond to reviewers' comments

16S rRNA gene sequencing is a powerful technique for describing and
comparing microbial communities (1). Efforts to link 16S rRNA gene
sequences to taxonomic levels based on distance thresholds date to at
least the 1990s. The distance-based threshold that was developed and is
now widely used was based on DNA-DNA hybridization approaches that are
not as precise as genome sequencing (2, 3). Instead, genome sequencing
technologies have suggested that the widely used 3\% distance threshold
to operationally define bacterial taxa is too coarse (4--6). As an
alternative to operational taxonomic units (OTUs), amplicon sequencing
variants (ASVs) have been proposed as a way to adopt the thresholds
suggested by genome sequencing to microbial community analysis using 16S
rRNA gene sequences (7--10). Approaches for identifying ASVs do not
cluster sequences based on a distance-based threshold (11). Proponents
of ASVs are largely dissmissive of concerns that most bacterial genomes
have more than one copy of the \emph{rrn} operon and that those copies
are not identical (12, 13). Yet, ASVs and using too fine a threshold to
identify OTUs could split a single genome into multiple clusters.
Conversely, using too broad of a threshold to define OTUs could cluster
together multiple bacterial species into the same OTU. An example of
both is seen in the comparison of \emph{Staphylococcus aureus} (NCTC
8325) and \emph{S. epidermidis} (ATCC 12228) where each genome has 5
copies of the 16S rRNA gene. Each of the 10 copies of the 16S rRNA gene
in these two genomes is distinct and represent 10 ASVs. Conversely, if
the copies were clustered using a 3\% distance threshold, then all 10
ASVs would cluster into the same OTU. The goal of this study was to
quantify the tradeoff of splitting a single genome into multiple
clusters and the risk of clustering different bacterial species into the
same cluster when using ASVs and various OTU definitions.

@@

At broad thresholds, 16S rRNA gene sequences from multiple species could
be clustered into the same ASV or OTU. I again randomly selected one
genome from each species to control for uneven representation of genomes
across species and for this analysis I measured the percentage of ASVs
and OTUs that contained 16S rRNA gene sequences from multiple species
(Figure 2). Without using distance-based thresholds, 4.1\% of the ASVs
contained sequences from multiple species when considering full length
sequences and 10.9, 16.2, and 13.1\% when considering the V3-V4, V4, and
V4-V5 regions, respectively. At the commonly used 3\% threshold for
defining OTUs, 27.4\% of the OTUs contained 16S rRNA gene sequences from
multiple species when considering full length sequences and 31.7, 34.3,
and 34.8\% when considering the V3-V4, V4, and V4-V5 regions,
respectively. Considering that species designations are inconsistently
applied and reflect multiple human-imposed biases, the risk of splitting
a genome into multiple OTUs is more problematic than clustering species
together. Therefore, larger thresholds are advisable.

The results of this analysis demonstrate that there is a significant
risk of splitting a single genome into multiple clusters if using ASVs
or too fine of a threshold to define OTUs. An ongoing problem for
amplicon-based studies is defining a meaningful taxonomic unit (11, 15,
16). Since there is no consensus for a biologicaly definition of a
bacterial species (17--19), microbiologists must accept that how
bacterial species are named is biased and that taxonomic rules are not
applied in a consistent manner (e.g., (19, 20)). This makes it
impossible to fit a distance threshold to define an OTU definition that
matches a set of species names (21). Furthermore, the 16S rRNA gene does
not evolve at the same rate across all bacterial lineages (15), which
limits the biological interpretation of a common OTU definition. A
distance-based definition of a taxonomic unit based on 16S rRNA gene or
full genome sequences is, at best, operational and not grounded in
biological theory (15, 22--24). There is general agreement in bacterial
systematics that to classify an organism to a bacterial species,
phenotypic and genome sequence data are needed (17--20). A short
sequence from a bacterial genome simply cannot differentiate between
species. Moreover, it is difficult to defend a clustering threshold that
would split a single genome into multiple taxonomic units. It is not
biologically plausible to entertain the possibility that different
\emph{rrn} operons from the same genome would have different ecologies.
Although there are multiple reasons that proponents favor ASVs, the
significant risk of artificially splitting genomes into separate
clusters is too high to warrant their use.

@@

\textbf{(v) Reproducible data analysis.} The code to perform the
analysis in this manuscript and its history are available as a git-based
version control repository on GitHub
(\url{https://github.com/SchlossLab/Schloss_rrnAnalysis_mSphere_2021}).
The analysis can be regenerated using a GNU Make-based workflow that
made use of built-in bash tools (v. 3.2.57), mothur (v. 1.44.2), and R
(v. 4.0.4). Within R, I used the tidyverse (v. 1.3.0), data.table (v.
1.13.2), Rcpp (v. 1.0.5), furrr (v. 0.2.1), here (v. 1.0.1) and
rmarkdown (v. 2.5) packages. The conception and development of this
analysis is available as a playlist on the Riffomonas YouTube channel
(\url{https://youtube.com/playlist?list=PLmNrK_nkqBpL7m_tyWdQgdyurerttCsPY}).

@@

\leavevmode\hypertarget{ref-Callahan2017}{}%
11. Callahan BJ, McMurdie PJ, Holmes SP. 2017. Exact sequence variants
should replace operational taxonomic units in marker-gene data analysis.
\emph{The ISME Journal} 11:2639--2643. doi:
\href{https://doi.org/10.1038/ismej.2017.119}{10.1038/ismej.2017.119}.

\leavevmode\hypertarget{ref-Pei2010}{}%
12. Pei AY, Oberdorf WE, Nossa CW, Agarwal A, Chokshi P, Gerz EA, Jin Z,
Lee P, Yang L, Poles M, Brown SM, Sotero S, DeSantis T, Brodie E, Nelson
K, Pei Z. 2010. Diversity of 16S rRNA genes within individual
prokaryotic genomes. \emph{Applied and Environmental Microbiology}
76:3886--3897. doi:
\href{https://doi.org/10.1128/aem.02953-09}{10.1128/aem.02953-09}.

\leavevmode\hypertarget{ref-Sun2013}{}%
13. Sun D-L, Jiang X, Wu QL, Zhou N-Y. 2013. Intragenomic heterogeneity
of 16S rRNA genes causes overestimation of prokaryotic diversity.
\emph{Applied and Environmental Microbiology} 79:5962--5969. doi:
\href{https://doi.org/10.1128/aem.01282-13}{10.1128/aem.01282-13}.

>>>>

16S rRNA gene sequencing is a powerful technique for describing and
comparing microbial communities (1). Efforts to link 16S rRNA gene
sequences to taxonomic levels based on distance thresholds date to at
least the 1990s. The distance-based threshold that was developed and is
now widely used was based on DNA-DNA hybridization approaches that are
not as precise as genome sequencing (2, 3). Instead, genome sequencing
technologies have suggested that the widely used 3\% distance threshold
to operationally define bacterial taxa is too coarse (4--6). As an
alternative to operational taxonomic units (OTUs), amplicon sequencing
variants (ASVs) have been proposed as a way to adopt the thresholds
suggested by genome sequencing to microbial community analysis using 16S
rRNA gene sequences (7--10). It is widely understood that individual
bacterial genomes often have multiple 16S rRNA genes that are not
identical and that a 16S rRNA gene sequence could be found with
different versions of the sequence in different genomes (11, 12). This
could lead to the problem that ASVs and using too fine a threshold to
identify OTUs could split a single genome into multiple clusters.
Proponents of ASVs minimize concerns that most bacterial genomes have
more than one copy of the \emph{rrn} operon and that those copies are
not identical (6, 13). Conversely, using too broad of a threshold to
define OTUs could cluster together multiple bacterial species into the
same OTU. An example of both is seen in the comparison of
\emph{Staphylococcus aureus} (NCTC 8325) and \emph{S. epidermidis} (ATCC
12228) where each genome has 5 copies of the 16S rRNA gene. Each of the
10 copies of the 16S rRNA gene in these two genomes is distinct and
represent 10 ASVs. Conversely, if the copies were clustered using a 3\%
distance threshold, then all 10 ASVs would cluster into the same OTU.
The goal of this study was to quantify the tradeoff of splitting a
single genome into multiple clusters and the risk of clustering
different bacterial species into the same cluster when using ASVs and
various OTU definitions.

@@

At broad thresholds, 16S rRNA gene sequences from multiple species could
be clustered into the same ASV or OTU. I again randomly selected one
genome from each species to control for uneven representation of genomes
across species. For this analysis I measured the percentage of ASVs and
OTUs that contained 16S rRNA gene sequences from multiple species
(Figure 2). Without using distance-based thresholds, 4.1\% of the ASVs
contained sequences from multiple species when considering full length
sequences and 10.9, 16.2, and 13.1\% when considering the V3-V4, V4, and
V4-V5 regions, respectively. At the commonly used 3\% threshold for
defining OTUs, 27.4\% of the OTUs contained 16S rRNA gene sequences from
multiple species when considering full length sequences and 31.7, 34.3,
and 34.8\% when considering the V3-V4, V4, and V4-V5 regions,
respectively. Although the actual fraction of ASVs and OTUs that contain
sequences from multiple species is dependent on the taxonomic
composition of the sequences in the \emph{rrn}DB, this analysis
highlights the tradeoffs of using distance-based thresholds.

The results of this analysis demonstrate that there is a significant
risk of splitting a single genome into multiple clusters if using ASVs
or too fine of a threshold to define OTUs. An ongoing problem for
amplicon-based studies is defining a meaningful taxonomic unit (13, 15,
16). Since there is no consensus for a biologicaly definition of a
bacterial species (17--19), microbiologists must accept that how
bacterial species are named is biased and that taxonomic rules are not
applied in a consistent manner (e.g., (19, 20)). This makes it
impossible to fit a distance threshold to define an OTU definition that
matches a set of species names (21). Furthermore, the 16S rRNA gene does
not evolve at the same rate across all bacterial lineages (15), which
limits the biological interpretation of a common OTU definition. A
distance-based definition of a taxonomic unit based on 16S rRNA gene or
full genome sequences is operational and not necessarily grounded in
biological theory (15, 22--24). One benefit of a distance-based OTU
definition is the ability to mask residual sequencing error. The
analysis in this study was conducted using ideal sequences from
assembled genomes whereas sequences generated in microbiome studies
would harbor PCR and sequencing errors. These errors would only
exacerbate the inflated number of ASVs. There is general agreement in
bacterial systematics that to classify an organism to a bacterial
species, phenotypic and genome sequence data are needed (17--20). A
short sequence from a bacterial genome simply cannot differentiate
between species. Moreover, it is difficult to defend a clustering
threshold that would split a single genome into multiple taxonomic
units. It is not biologically plausible to entertain the possibility
that different \emph{rrn} operons from the same genome would have
different ecologies. Individual bacteria are defined at the cellular or
chromosomal level and not at the gene level. One could argue that, in
practice, communities are compared on a relative rather than absolute
basis. However, communities harboring populations that tend to have more
copies of the \emph{rrn} operon would appear to have higher richness and
diversity than those with fewer copies purely due to the propensity for
populations with more \emph{rrn} operons to generate more ASVs. Although
there are multiple reasons that proponents favor ASVs, the significant
risk of artificially splitting genomes into separate clusters is too
high to warrant their use.

@@

\textbf{(v) Reproducible data analysis.} The code to perform the
analysis in this manuscript and its history are available as a git-based
version control repository on GitHub
(\url{https://github.com/SchlossLab/Schloss_rrnAnalysis_mSphere_2021}).
The analysis can be regenerated using a GNU Make-based workflow that
made use of built-in bash tools (v. 3.2.57), mothur (v. 1.44.2), and R
(v. 4.1.0). Within R, I used the tidyverse (v. 1.3.1), data.table (v.
1.14.0), Rcpp (v. 1.0.6), furrr (v. 0.2.2), here (v. 1.0.1) and
rmarkdown (v. 2.8) packages. The conception and development of this
analysis is available as a playlist on the Riffomonas YouTube channel
(\url{https://youtube.com/playlist?list=PLmNrK_nkqBpL7m_tyWdQgdyurerttCsPY}).

@@

\leavevmode\hypertarget{ref-Pei2010}{}%
11. Pei AY, Oberdorf WE, Nossa CW, Agarwal A, Chokshi P, Gerz EA, Jin Z,
Lee P, Yang L, Poles M, Brown SM, Sotero S, DeSantis T, Brodie E, Nelson
K, Pei Z. 2010. Diversity of 16S rRNA genes within individual
prokaryotic genomes. \emph{Applied and Environmental Microbiology}
76:3886--3897. doi:
\href{https://doi.org/10.1128/aem.02953-09}{10.1128/aem.02953-09}.

\leavevmode\hypertarget{ref-Sun2013}{}%
12. Sun D-L, Jiang X, Wu QL, Zhou N-Y. 2013. Intragenomic heterogeneity
of 16S rRNA genes causes overestimation of prokaryotic diversity.
\emph{Applied and Environmental Microbiology} 79:5962--5969. doi:
\href{https://doi.org/10.1128/aem.01282-13}{10.1128/aem.01282-13}.

\leavevmode\hypertarget{ref-Callahan2017}{}%
13. Callahan BJ, McMurdie PJ, Holmes SP. 2017. Exact sequence variants
should replace operational taxonomic units in marker-gene data analysis.
\emph{The ISME Journal} 11:2639--2643. doi:
\href{https://doi.org/10.1038/ismej.2017.119}{10.1038/ismej.2017.119}.
